Predicting I/O-performance in HPC using Artificial Neural Networks

### 2016-07-04 Paper/paper.tex
Bilder access sizes fertig
Aufbau des Papers, etc.

\keywords{file systems, performance, predicting file access times, artificial neural networks}
\end{abstract}

@@

Tools are demanded that help users of HPC-facilities to implement efficient input/output (I/O) in their programs.
It is difficult to find the best access parameters and patterns due to complex parallel storage systems.
The processing of file accesses in a storage system can be viewed as a task that is sequentially propagated along a I/O-path in the storage system.
Starting at the invoking processor, the storage system is searching for the data, going further and further through the storage hierarchy, until all data is found, so it can be returned to the processor.

@@

This paper is organized as follows:
Related work is provided in Section\,\ref{sec:related}.
% TODO error class...
Firstly, we analyze behavior of the storage system by measuring a series of access times.
Then, a simplified model of the I/O path is developed.
Next, we describe how we derive potential classes (alternative I/O path) from the error of a prediction model.
Finally, a set of models

@@

\section{Modeling I/O Access Times}

\subsection{Characteristics of the Data}
This approach allows for an in-depth analysis of system behavior for different use cases.
We used a synthetic benchmark in which identical measurements for random or sequential I/O is performed and the access time is measured individually. % , instead of using arbitrary access parameters for consecutive accesses ??? TODO was meint das
The resulting timings are stored together with file access parameters in a file that is then used to build and validate the models.
The stored parameters are:
\begin{itemize}
%	\item \textbf{File-ID}: Identifies the invoked file. Brauchen wir ja nicht für die Analyse
	\item \textbf{Access size}: Number of bytes to read or write.
	\item \textbf{Access type}: Differentiates reading and writing file accesses.
\end{itemize}
Directly measurable or derivable attributes are:
\begin{itemize}
	\item \textbf{Offset}: Distance of file beginning to the starting point of access.
	\item \textbf{Delta-Offset}: Can be calculated for file access $i$ as Offset[$i$] - (Offset[i-1] + access size[i-1]).
	\item \textbf{Access time}: Time for performing the I/O.
\end{itemize}
Knowledge of internal aspects of the system about the current system utilization or about the storage media that have to be addressed is not directly available.

\subsection{Model of the I/O-path}
The internal processing of a file access in the storage system can be viewed using the I/O-path which is the path from the invoking processor to the storage medium that contains the data (larger accesses may be distributed across multiple servers). 
Once the data has been found it is passed-through the levels of storage hierarchy to caches.
The resulting access time depends on the depth of this I/O-path because storage media further along the path are increasingly slower.
(\textbf{figure of storage hierarchy?} TODO OK).
While the first levels in storage hierarchy (CPU caches) are extremely fast, the main memory is already slower; the same applies for the step into the parallel storage system, that is connected via network to the computer nodes.\\
Reading file accesses are more severely affected by varying depths of I/O-paths than writing file accesses, which are usually propagated to the disk drives.

@@

\subsection{Models}
Access times of file accesses can be predicted using various models.
While we evaluated several models for this paper, only a few relevant models are described; more models are described in \cite{VVEIHUDVVN15}. 
% TODO: Wo soll ich meine Arbeit noch zitieren?
Each model seeks to correlate its known file access parameters to the corresponding access time. 
Linear regression is used as a baseline model with a simple mapping of access size to access time.
Additionally, three models with different input information utilizing ANNs were used:
\begin{itemize}
	\item The most simple ANN-model only received information about access sizes, delta-offsets and access types as input.
	\item A second ANN-model, called \textbf{ema-model}, received additionally to the parameters of the previous model information about the past data throughputs of the system, which can be used to exploit time dependencies of the I/O-performance.
	\item The third ANN-model, called \textbf{error-class-model}, received error classes in addition to the access parameters of the first ANN-model.
\end{itemize}\medskip

The idea of the ema-model is to exploit periodic changes in the performance of the storage system.
Our analysis of measurements as well as the work of Crume et al. \cite{Crume:2013:FML:2538542.2538561} indicate that periodic phenomenons of access times in sequences of files accesses can be exploited for access time prediction.\\
Figure \ref{src/plot_From100001to100250_read_seq.png} shows a small series of measurements of sequentially read file accesses.
The peaks in access time are occurring periodically in this particular sequence. Such behavior appears only partially, often access times fluctuate without observable periodicity as in Figure.
An occurrence of a periodic pattern could be explained with a processing of the storage system in which a larger amount of data is loaded into the cache at the same time as soon as a cache-miss happens.
\fig{width=0.47\textwidth}{src/plot_From100001to100250_read_seq.png}{Small sequence of reading file accesses with periodic peaks in access time.}
\fig{width=0.47\textwidth}{src/plot_First250_read_seq.png}{Small sequence of reading file accesses with sporadic peaks in access time.}
With knowledge about these period of peaks a model could make better access time predictions.
Our second model is supposed to use its input information of previous data throughput for that.
The data throughput has peaks the same period as the access time.
Internally the model calculates the exponential moving average (EMA) of data throughput of measurement $i$ as:
\begin{equation}
EMA(i) = 0.5 \cdot t(i)+ 0.5 \cdot EMA(i-1)
\end{equation}
With $t(i)$ the access time of measurement $i$.
The influence of data throughput of a measurement is exponentially decreasing in the series of EMAs.
In Figure \ref{src/ema2.png} the EMA-function to a graph with periodic peaks can be seen.
\fig{width=.47\textwidth}{src/ema2.png}{Exponential moving average in red for a graph with perdiodic peaks.}\\
The ema-model can use $EMA(n-1)$ for the prediction of access time of measurement $n$.
If the model is able to find threshold values of the EMA-function after which another peaks follows, it can use this input information for better access time predictions on sequences with periodic peaks in access time. Depending on how relevant periodic behavior of access times is in our measurements, this additional input information could improve the model.

%The error-class-model is used as described in Section\,\ref{sec:error_classes} to examine the potential improvements for access time prediction due to knowledge about I/O-paths. TODO: wofür nochmal ?
% TODO: Soll hier nicht nochmal erwähnt werden, wofür das error-class-model genutzt wird?

\section{Evaluation}
The execution and results of analysis are summarized in the following.
First the test system used for measurements and the strategy for systematic measurement are presented.
Then follows the study of received data and computed error classes.
And in the end the quality of predictions of the different models are examined.

@@

% TODO: können wir die Datengröße noch dran schreiben? Einfach mit Gimp die Größe vertical drunter fügen anstelle Index ...
% TODO: Global fixen.
% TODO Kann man in deren Style auch zwei BIlder nebeneinander zeichnen? Das wäre Cool. Ansonstne OK...

@@

\section{Conclusion and future work}
In this work, we analyzed the performance of a supercomputer's storage system.
Using a machine learning approach with artificial neural networks, we developed different models for file access time prediction.
Through our study of measured file access times and model results we were able to gain knowledge about the behavior of the storage system.
We found out that linear models are not feasible for access time prediction, despite the strong correlation of access time to access size.
Our models utilizing ANNs achieved much better results than linear regression.\\
The hypothesis that knowledge about the internal processing of a file access in form of I/O-paths is essential for access time prediction was supported by our data.
This is because file accesses with equal access parameters can produce strongly deviating access times depending on the used I/O-path.\\
Our approach of deriving knowledge about I/O-paths by exploiting periodic performance of the storage system was not successful.
The additional input information did not lead to a significant improvement of access time predictions.\\
However, with our method of clustering residues of linear regression for approximations of I/O-paths as error classes, we were able to illustrate the importance of I/O-paths for access time prediction. The ANN-model with additional input information of error classes was able to reduce its average prediction error to a third compared to the ANN-model with only access parameter values as input information.\medskip

>>>>

\keywords{file systems, performance, predicting file access times, artificial neural networks}%TODO Gute keywords?
\end{abstract}

@@

Tools are demanded that help users of HPC-facilities to implement efficient input/output (I/O) in their programs.
It is difficult to find the best access parameters and patterns due to complex parallel storage systems.
The processing of file accesses in a storage system can be viewed as a task that is sequentially propagated along a I/O-path in the storage system.
Starting at the invoking processor, the storage system is searching for the data, going further and further through the memory hierarchy, until all data is found, so it can be returned to the processor.

@@

This paper is organized as follows:
Related work is provided in Section\,\ref{sec:related}.
In Section\,\ref{modeling_access_times}, we explain our analysis of file access times. 
First we develop a simplified model of the I/O-path. Then a method for approximating I/O-paths with derived classes from the error of a prediction model is proposed. At last we introduce a set of models for access time prediction.
Afterwards in Section\,\ref{evaluation}, we evaluate our analysis of the storage system by measuring access times in various test series, studying the obtained access times and the predictions of our models to them.
In the end Section\,\ref{conclusion} summarizes the paper suggests possible future work.

@@

\section{Modeling I/O Access Times}
\label{modeling_access_times}
\subsection{Characteristics of the Data}
This approach allows for an in-depth analysis of system behavior for different use cases.
We used a synthetic benchmark in which identical measurements for random or sequential I/O are performed and the access time is measured individually.
The resulting timings are stored together with file access parameters in a file that is then used to build and validate the models.
The stored parameters are:
\begin{itemize}
%	\item \textbf{File-ID}: Identifies the invoked file. Brauchen wir ja nicht für die Analyse
	\item \textbf{Access size}: Number of bytes to read or write.
	\item \textbf{Access type}: Differentiates reading and writing file accesses.
\end{itemize}
Directly measurable or derivable attributes are:
\begin{itemize}
	\item \textbf{Offset}: Distance of file beginning to the starting point of access.
	\item \textbf{Delta-Offset}: Can be calculated for file access $i$ as Offset[$i$] - (Offset[i-1] + access size[i-1]).
	\item \textbf{Access time}: Time for performing the I/O.
\end{itemize}
Knowledge of internal aspects of the system about the current system utilization or about the storage media that have to be addressed is not directly available.

\subsection{Model of the I/O-path}
The internal processing of a file access in the storage system can be viewed using the I/O-path which is the path from the invoking processor to the storage medium that contains the data (larger accesses may be distributed across multiple servers). 
Once the data has been found it is passed-through the levels of memory hierarchy to caches.
The resulting access time depends on the depth of this I/O-path because storage media further along the path are increasingly slower.
While the first levels in memory hierarchy (CPU caches) are extremely fast, the main memory is already slower; the same applies for the step into the parallel storage system, that is connected via network to the computer nodes.
%The trade-off between memory capacity and performance can be visualized with the memory hierarchy as in figure %\ref{src/2000px-ComputerMemoryHierarchy.png}.
%\fig{width=.7\textwidth}{src/2000px-ComputerMemoryHierarchy.png}{Memory hierarchy.}
Reading file accesses are more severely affected by varying depths of I/O-paths than writing file accesses, which are usually propagated to the disk drives.

@@

\subsection{Models}
Access times of file accesses can be predicted using various models.
While we evaluated several models for this paper, only a few relevant models are described; more models are described in \cite{VVEIHUDVVN15}. 
% TODO: Soll ich meine Arbeit noch woanders zitieren?
Each model seeks to correlate its known file access parameters to the corresponding access time. 
Linear regression is used as a baseline model with a simple mapping of access size to access time.
Additionally, three models with different input information utilizing ANNs were used:
\begin{itemize}
	\item The most simple ANN-model only received information about access sizes, delta-offsets and access types as input.
	\item A second ANN-model, called \textbf{ema-model}, received additionally to the parameters of the previous model information about the past data throughputs of the system, which can be used to exploit time dependencies of the I/O-performance.
	\item The third ANN-model, called \textbf{error-class-model}, received error classes in addition to the access parameters of the first ANN-model.
\end{itemize}\medskip

The idea of the ema-model is to exploit periodic changes in the performance of the storage system.
Our analysis of measurements as well as the work of Crume et al. \cite{Crume:2013:FML:2538542.2538561} indicate that periodic phenomenons of access times in sequences of files accesses can be exploited for access time prediction.\\
Figure \ref{src/plot_From100001to100250_read_seq.png} shows a small series of measurements of sequentially read file accesses.
The peaks in access time are occurring periodically in this particular sequence. Such behavior appears only partially, often access times fluctuate without observable periodicity as in Figure.
An occurrence of a periodic pattern could be explained with a processing of the storage system in which a larger amount of data is loaded into the cache at the same time as soon as a cache-miss happens.
\fig{width=0.47\textwidth}{src/plot_From100001to100250_read_seq.png}{Small sequence of reading file accesses with periodic peaks in access time.}
\fig{width=0.47\textwidth}{src/plot_First250_read_seq.png}{Small sequence of reading file accesses with sporadic peaks in access time.}
With knowledge about these period of peaks a model could make better access time predictions.
Our second model is supposed to use its input information of previous data throughput for that.
The data throughput has peaks the same period as the access time.
Internally the model calculates the exponential moving average (EMA) of data throughput of measurement $i$ as:
\begin{equation}
EMA(i) = 0.5 \cdot t(i)+ 0.5 \cdot EMA(i-1)
\end{equation}
With $t(i)$ the access time of measurement $i$.
The influence of data throughput of a measurement is exponentially decreasing in the series of EMAs.
In Figure \ref{src/ema2.png} the EMA-function to a graph with periodic peaks can be seen.
\fig{width=.47\textwidth}{src/ema2.png}{Exponential moving average in red for a graph with perdiodic peaks.}\\
The ema-model can use $EMA(n-1)$ for the prediction of access time of measurement $n$.
If the model is able to find threshold values of the EMA-function after which another peaks follows, it can use this input information for better access time predictions on sequences with periodic peaks in access time. Depending on how relevant periodic behavior of access times is in our measurements, this additional input information could improve the model.\medskip

The error-class-model is used as described in Section\,\ref{sec:error_classes} to examine the potential improvements for access time prediction due to knowledge about I/O-paths.
Therefore, we evaluate how much the prediction of access times of a model using error classes improves compared to a model without.

\section{Evaluation}
\label{evaluation}
The execution and results of analysis are summarized in the following.
First the test system used for measurements and the strategy for systematic measurement are presented.
Then follows the study of received data and computed error classes.
And in the end the quality of predictions of the different models are examined.

@@

% TODO Kann man in deren Style auch zwei BIlder nebeneinander zeichnen? Das wäre Cool. Das könnte ich probieren, aber kann man nicht auf irgendwie zwei nebeneinander setzen?

@@

\section{Conclusion and future work}
\label{conclusion}
In this work, we analyzed the performance of a supercomputer's storage system.
Using a machine learning approach with artificial neural networks, we developed different models for file access time prediction.
Through our study of measured file access times and model results we were able to gain knowledge about the behavior of the storage system.
We found out that linear models are not feasible for access time prediction, despite the strong correlation of access time to access size.
Our models utilizing ANNs achieved much better results than linear regression.\\
The hypothesis that knowledge about the internal processing of a file access in form of I/O-paths is essential for access time prediction was supported by our data.
This is because file accesses with equal access parameters can produce strongly deviating access times depending on the used I/O-path.\\
Our approach of deriving knowledge about I/O-paths by exploiting periodic performance of the storage system was not successful.
The additional input information did not lead to a significant improvement of access time predictions.\\
However, with our method of clustering residues of linear regression for approximations of I/O-paths as error classes, we were able to illustrate the importance of I/O-paths for access time prediction. The ANN-model with additional input information of error classes was able to reduce its average prediction error to a third compared to the ANN-model with only access parameter values as input information.\medskip
